\bf Effects without a Cause: The Search for Demographic Anomalies

### 2020-01-10 MANUSCRIPT/manuscript.tex
Update manuscript.tex

\bigskip
\begin{abstract}
The proliferation of data, modern computing advances, and powerful
statistical algorithms unlock the potential search for hidden,
previously un- or understudied demographic anomalies. Here, we
investigate US state-level fertility and mortality time series since
1999 to uncover the hidden baby booms/busts and mortality
plagues?/non-plauges?. We find 38 states exhibited at least one
mortality anomaly, totalling more than 318k anomalous deaths and an
additional 6 states exhibited at least one mortality non-plague?
totalling 164k protective deaths. 12 states exhibited baby busts,
totalling more than 240k missing births and an additional 11 states
exhibited baby booms totalling more than 134k additional births. These
results suggest the widespread detection of demographic anomalies. Our
analysis does not examine the \emph{causes} of these anomalies and our
results point to important further research on the causes of anomalous
demographic behavior.\\
\end{abstract}

\noindent%
{\it Keywords:} mortality, fertility, causal inference
\vfill

@@

Causal inference approaches in social sciences have a long history
\citep{grimmer2015ppsp}. Often, the identification of a casual mechanism
requires either a randomized control trial (RCT) or expert knowledge
applied to a natural experiment. For population-level analysis,
expensive RCTs usually preclude their widespread adoption
\citep{west2008ajph}. Using natural experiments, however, provides
important and more widespread insights on the determinants of migration
after Hurricane Katrina in 2006
\citep{fussellRecoveryMigrationCity2014, horiDisplacementDynamicsSouthern2009},
mini baby booms after electrical blackouts \citep{fetzer2018jpe}, and
the highly publicized estimates of excess mortality after Hurricane
Maria in Puerto Rico in 2016
\citep{kishore2018mortality, santos2018use}. Both RCTs and natural
experiments follow traditional, hypothesis testing, inductive scientific
paradigms where a question is first posed and scientists generate or
find data to answer the question. Such approaches, while extensively
utilized and time-tested, are likely to miss important phenomena that
might go unnoticed.

>>>>

\bigskip
\begin{abstract}
\noindent%
The proliferation of data, modern computing advances, and powerful
statistical algorithms unlock the potential search for hidden,
previously un- or understudied demographic anomalies. Here, we demonstrate how demographers can use causal inference techniques to identify anamolies by 
investigating US state-level fertility and mortality time series since
1999 to uncover the hidden baby booms/busts and mortality
plagues?/non-plauges?. We find 38 states exhibited at least one
mortality anomaly, totalling more than 318k anomalous deaths and an
additional 6 states exhibited at least one mortality non-plague?
totalling 164k protective deaths. 12 states exhibited baby busts,
totalling more than 240k missing births and an additional 11 states
exhibited baby booms totalling more than 134k additional births. These
results suggest the widespread detection of demographic anomalies. Our
analysis does not examine the \emph{causes} of these anomalies but our
results point to important further research on the causes of anomalous
demographic behavior.\\
\end{abstract}

\noindent%
{\it Keywords:} mortality, fertility, causal inference, population trends
\vfill

@@

Causal inference, in its most common usage, is an attempt to undercover the underlying mechanism that results in changes in a phenomenon. Causal inference approaches in social sciences have a long history
\citep{grimmer2015ppsp}. Often, the identification of a casual mechanism
requires either a randomized control trial (RCT) or expert knowledge
applied to a natural experiment. For population-level analysis,
expensive RCTs usually preclude their widespread adoption
\citep{west2008ajph}. Natural experiments provides less certainty about causation and fewer opportunities to replicate work; natural experiments provide
important and widespread insights such as the determinants of migration
after Hurricane Katrina in 2006
\citep{fussellRecoveryMigrationCity2014, horiDisplacementDynamicsSouthern2009},
mini baby booms after electrical blackouts \citep{fetzer2018jpe}, and
the highly publicized estimates of excess mortality after Hurricane
Maria in Puerto Rico in 2016
\citep{kishore2018mortality, santos2018use}.

Both RCTs and natural
experiments follow traditional hypothesis testing--inductive scientific
paradigms where a question is first posed and scientists generate or
find data to answer the question. Such approaches, while extensively
utilized and time-tested, are likely to miss important phenomena that
might go unnoticed. Data scientists are rethinking these approaches and have re-envisioned causal inference as a machine learning technique that allows big data researchers to move from correlations that have been established with good accuracy to making strong conclusions about the underlying mechanisms that produce these correlations
\citep{Pearl2018}.
